Do investors in clean energy ETFs herd?

### 2025-01-03 alt_energy_herding.qmd
complete introduction and literature review.

# Introduction

Climate sustainability has become a major concern for financial markets and investors. This growing interest is primary driven by the increasing materialisation of climate risks, and the actions of governments, institutions and organizations towards a sustainable future [@giglio2021a]. In order to ensure returns investors increasingly seek to herd against climate risks by investing in green financial products. Although the evidence is mixed, indications that returns from green financial products are comparable to traditional financial [see amongst other, @decclesia2024; @nguyen2025; @pastor2022; and @naqvi2022]. Therefore, the growth of green assets-under-management is likely to continue.

A key feature of the green financial products is the exchange traded funds (ETFs). ETFs are a type of security that involves a collection of securities that often tracks an underlying index, although they can invest in any number of industry sectors or use various strategies. In addition, environment, social, and governance (ESG) ETFs serve as a market discovery tool for investors to identify and invest in companies that are environmentally friendly [@briere2023]. Thus, ESG ETFs exhibit environment friendly characteristics.

Among the ESG ETFs, the Clean Energy (CE) ETFs have been the best-performing ones in 2022, followed by the Cybersecurity and Artificial Intelligence (AI) ETFs [@decclesia2024]. The clean energy transition represents one of the largest multi-decade secular growth opportunities.  After the inclusion of Green energy financing in the list of United Nations Sustainability Goals (SDGs) as SDG 7, the role, importance, and visibility of green financial products; all have escalated enormously [@naqvi2022].

>>>>

# Introduction and literature review

Climate sustainability is a major concern for financial markets and investors. The growing interest in climate sustainability is primarily driven by the increasing materialisation of climate risks, and the actions of governments, institutions and organizations towards a sustainable future [@giglio2021a]. To ensure returns, investors increasingly seek to hedge against climate risks by investing in green financial products. Although the evidence is mixed, indications that returns from green financial products are comparable to traditional financial [see amongst others, @decclesia2024; @nguyen2025; @pastor2022; and @naqvi2022].

In this new climate sustainability paradigm, investors face many pressures that not only bear on returns but also the stability of financial markets. For example, resulting regulations aimed at reducing emissions can surprisingly reduce the profitability of fossil-fuel-based companies, or the possible mispricing of assets from ignoring climate risks can lead to significant losses [@nguyen2025]. In addition to these climate risks, a general change in investor attitudes can drive the inclusion of green assets in their portfolio leading to systemic risk.^[These pressures notwithstanding the possible contribution that financial markets can play in mitigating and reducing the negative effects of climate change [@giglio2021a].

Specifically, climate change presents risks to investor portfolios through two primary sources - physical and transition risks. Physical risks or direct impact refer to extreme climate events such as floods and droughts, which impact business operations and infrastructure; and transition risks are the policy, technological, and other costs that societies bear to achieve low carbon economies [@nguyen2025; and @giglio2021a amongst others]. Investors, therefore, recognise these risks and seek to mitigate them, as they seek return-enhancing green financial products.

Exchange-traded funds (ETFs) are a key feature of green financial products. ETFs are a type of security that involves a collection of securities that often tracks an underlying index. However, they can invest in various industry sectors or strategies. In addition, environment, social, and governance (ESG) ETFs serve as a market discovery tool for investors to identify and invest in environmentally friendly companies [@briere2023]. Among the ESG ETFs, the Clean Energy (CE) ETFs have been the best-performing ones in 2022, followed by the Cybersecurity and Artificial Intelligence (AI) ETFs [@decclesia2024]. The clean energy transition represents one of the largest multi-decade secular growth opportunities.  After the inclusion of Green energy financing in the list of United Nations Sustainability Goals (SDGs) as SDG 7, the role, importance, and visibility of green financial products; all have escalated enormously [@naqvi2022]. Therefore, the growth of green assets under management is likely to continue.

However, given how recent the inclusion of climate sustainability investment decisions is, it is not clear what the actual impact will be in the long run. In this study specifically, we ask whether the rapid adoption of CE ETFs could be driven by market fads, or is a fundamental change in investor behaviour. Investors, for example, can believe that peers have information about climate risks that they do not, investors may herd to avoid losses compared to peers, or investors may be encouraged to herd by the desire to align to climate-related social values [@ciciretti2021; @gavrilakis2023; @loang2023]. Therefore, market volatility and crisis, financial performance, and investor sentiment can drive herding in ESG markets.

Herding behaviour in ESG markets is not without precedent. Amongst others, @loang2023 found that compliance with SDG goals can introduce bias in investor sentiment, which leads to herding behaviour. Using a Twitter (or X) uncertainty index, @koutmos2024 found evidence of herding in US-based ESG index fund investors. @przychodzen2016 found herding behaviour amongst fund managers who incorporated ESG strategies in their portfolios. Lastly, @rubbaniy2021 highlight evidence of herding in the MSCI US ESG Leader Index during extreme periods (bear and bull periods).

However, to the best of our knowledge, no studies exist that focus on herding behaviour in CE ESGs. This study, therefore, aims to extend the broader literature on herding behaviour in commodity and energy markets. Several studies in this area were conducted. For example, @demirer2013a conducted a sector study and found herding behaviour in grains but not in other sectors. Similarly, @gilbert2010 showed herding behaviour amongst speculators in non-ferrous commodities. Others did not find evidence of herding in similar markets. @babalos2015 found significant anti-herding behaviour in metal commodities futures after the global financial crisis. @pierdzioch2010 showed that forecasters in oil and metals markets deviated from the crowd, indicating a rational response to information. @steen2013 also found no herding behaviour in international commodity markets. Overall, the literature in this area is mixed, which indicates scope for further study.

Therefore,  this study investigates herding behaviour in alternative energy ETFs in the US between 2019 and 2024. We then demonstrate how climate-related uncertainty can drive herding behaviour in these markets. Methodologically, we follow the standard herding tests by @christie1995 and @chang2000. The traditional approach was supplemented by quantile regressions [@koenker1978 ] in order to capture the time-varying aspects of herding. Lastly,  we extend recent approaches by @bua2024 and others, which seek to establish a link between climate uncertainty and herding behaviour. The next section describes the data and methodology, followed by the results and conclusions.

### 2025-01-03 alt_energy_herding.tex
complete introduction and literature review.

\setstretch{1.5}
\section{Introduction}\label{introduction}

Climate sustainability has become a major concern for financial markets
and investors. This growing interest is primary driven by the increasing
materialisation of climate risks, and the actions of governments,
institutions and organizations towards a sustainable future
\citep{giglio2021a}. In order to ensure returns investors increasingly
seek to herd against climate risks by investing in green financial
products. Although the evidence is mixed, indications that returns from
green financial products are comparable to traditional financial
\citetext{\citealp[see amongst
other,][]{decclesia2024}; \citealp{nguyen2025}; \citealp{pastor2022}; \citealp[and][]{naqvi2022}}.
Therefore, the growth of green assets-under-management is likely to
continue.

A key feature of the green financial products is the exchange traded
funds (ETFs). ETFs are a type of security that involves a collection of
securities that often tracks an underlying index, although they can
invest in any number of industry sectors or use various strategies. In
addition, environment, social, and governance (ESG) ETFs serve as a
market discovery tool for investors to identify and invest in companies
that are environmentally friendly \citep{briere2023}. Thus, ESG ETFs
exhibit environment friendly characteristics.

Among the ESG ETFs, the Clean Energy (CE) ETFs have been the
best-performing ones in 2022, followed by the Cybersecurity and
Artificial Intelligence (AI) ETFs \citep{decclesia2024}. The clean
energy transition represents one of the largest multi-decade secular
growth opportunities. After the inclusion of Green energy financing in
the list of United Nations Sustainability Goals (SDGs) as SDG 7, the
role, importance, and visibility of green financial products; all have
escalated enormously \citep{naqvi2022}.

>>>>

\setstretch{1.5}
\section{Introduction and literature
review}\label{introduction-and-literature-review}

Climate sustainability is a major concern for financial markets and
investors. The growing interest in climate sustainability is primarily
driven by the increasing materialisation of climate risks, and the
actions of governments, institutions and organizations towards a
sustainable future \citep{giglio2021a}. To ensure returns, investors
increasingly seek to hedge against climate risks by investing in green
financial products. Although the evidence is mixed, indications that
returns from green financial products are comparable to traditional
financial \citetext{\citealp[see amongst
others,][]{decclesia2024}; \citealp{nguyen2025}; \citealp{pastor2022}; \citealp[and][]{naqvi2022}}.

In this new climate sustainability paradigm, investors face many
pressures that not only bear on returns but also the stability of
financial markets. For example, resulting regulations aimed at reducing
emissions can surprisingly reduce the profitability of fossil-fuel-based
companies, or the possible mispricing of assets from ignoring climate
risks can lead to significant losses \citep{nguyen2025}. In addition to
these climate risks, a general change in investor attitudes can drive
the inclusion of green assets in their portfolio leading to systemic
risk.\^{}{[}These pressures notwithstanding the possible contribution
that financial markets can play in mitigating and reducing the negative
effects of climate change \citep{giglio2021a}.

Specifically, climate change presents risks to investor portfolios
through two primary sources - physical and transition risks. Physical
risks or direct impact refer to extreme climate events such as floods
and droughts, which impact business operations and infrastructure; and
transition risks are the policy, technological, and other costs that
societies bear to achieve low carbon economies
\citetext{\citealp{nguyen2025}; \citealp[and][ amongst
others]{giglio2021a}}. Investors, therefore, recognise these risks and
seek to mitigate them, as they seek return-enhancing green financial
products.

Exchange-traded funds (ETFs) are a key feature of green financial
products. ETFs are a type of security that involves a collection of
securities that often tracks an underlying index. However, they can
invest in various industry sectors or strategies. In addition,
environment, social, and governance (ESG) ETFs serve as a market
discovery tool for investors to identify and invest in environmentally
friendly companies \citep{briere2023}. Among the ESG ETFs, the Clean
Energy (CE) ETFs have been the best-performing ones in 2022, followed by
the Cybersecurity and Artificial Intelligence (AI) ETFs
\citep{decclesia2024}. The clean energy transition represents one of the
largest multi-decade secular growth opportunities. After the inclusion
of Green energy financing in the list of United Nations Sustainability
Goals (SDGs) as SDG 7, the role, importance, and visibility of green
financial products; all have escalated enormously \citep{naqvi2022}.
Therefore, the growth of green assets under management is likely to
continue.

However, given how recent the inclusion of climate sustainability
investment decisions is, it is not clear what the actual impact will be
in the long run. In this study specifically, we ask whether the rapid
adoption of CE ETFs could be driven by market fads, or is a fundamental
change in investor behaviour. Investors, for example, can believe that
peers have information about climate risks that they do not, investors
may herd to avoid losses compared to peers, or investors may be
encouraged to herd by the desire to align to climate-related social
values \citep{ciciretti2021, gavrilakis2023, loang2023}. Therefore,
market volatility and crisis, financial performance, and investor
sentiment can drive herding in ESG markets.

Herding behaviour in ESG markets is not without precedent. Amongst
others, \citet{loang2023} found that compliance with SDG goals can
introduce bias in investor sentiment, which leads to herding behaviour.
Using a Twitter (or X) uncertainty index, \citet{koutmos2024} found
evidence of herding in US-based ESG index fund investors.
\citet{przychodzen2016} found herding behaviour amongst fund managers
who incorporated ESG strategies in their portfolios. Lastly,
\citet{rubbaniy2021} highlight evidence of herding in the MSCI US ESG
Leader Index during extreme periods (bear and bull periods).

However, to the best of our knowledge, no studies exist that focus on
herding behaviour in CE ESGs. This study, therefore, aims to extend the
broader literature on herding behaviour in commodity and energy markets.
Several studies in this area were conducted. For example,
\citet{demirer2013a} conducted a sector study and found herding
behaviour in grains but not in other sectors. Similarly,
\citet{gilbert2010} showed herding behaviour amongst speculators in
non-ferrous commodities. Others did not find evidence of herding in
similar markets. \citet{babalos2015} found significant anti-herding
behaviour in metal commodities futures after the global financial
crisis. \citet{pierdzioch2010} showed that forecasters in oil and metals
markets deviated from the crowd, indicating a rational response to
information. \citet{steen2013} also found no herding behaviour in
international commodity markets. Overall, the literature in this area is
mixed, which indicates scope for further study.

Therefore, this study investigates herding behaviour in alternative
energy ETFs in the US between 2019 and 2024. We then demonstrate how
climate-related uncertainty can drive herding behaviour in these
markets. Methodologically, we follow the standard herding tests by
\citet{christie1995} and \citet{chang2000}. The traditional approach was
supplemented by quantile regressions \citep[ ]{koenker1978} in order to
capture the time-varying aspects of herding. Lastly, we extend recent
approaches by \citet{bua2024} and others, which seek to establish a link
between climate uncertainty and herding behaviour. The next section
describes the data and methodology, followed by the results and
conclusions.
